Estimating the number of clusters using Cross Validation

### 2015-06-19 cvclust.tex
minor cleanups to tex file

\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\newcommand{\dataX}{\mathfrak{X}}
\newcommand{\Xtrain}{X_{\text{train}}}
\newcommand{\Ytrain}{Y_{\text{train}}}
\newcommand{\Xtest}{X_{\text{test}}}
\newcommand{\Ytest}{Y_{\text{test}}}
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}
\newcommand{\bpi}{\bar \pi}
\newcommand{\bmuX}{\bar \mu^{X}}
\newcommand{\bmuY}{\bar \mu^{Y}}
\newcommand{\sA}{\mathcal{A}}
\newcommand{\sbA}{\mathcal{\bar A}}
\newcommand{\sC}{\mathcal{C}}
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sN}{\mathcal{N}}
\newcommand{\sX}{\mathcal{X}}

@@

Cross-validation is commonly used for model selection in supervised learning
problems.  In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\mathbb{R}^{p}$ and $Y_i \in \mathbb{R}^{q}$.  The data can be represented as
a matrix with $N$ rows and $p + q$ columns.  We partition the data into $K$
hold-out ``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
cross-validation, we permute the rows and columns of the data matrix and then
partition the rows and columns as $N = n + m$ and $P = p + q$ for 
non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
columns as ``predictors'' and the last $q$ columns as ``responses'';
similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
as ``test''.  In block form, the permuted data matrix is
\[
  \dataX
  =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix},
\]
where
$\Xtrain \in \mathbb{R}^{n \times p}$,
$\Ytrain \in \mathbb{R}^{n \times q}$,
$\Xtest \in \mathbb{R}^{m \times p}$,
and
$\Ytest \in \mathbb{R}^{m \times q}$.

Given such a partition of $\dataX$, we perform four steps for each value of
$k$, the number of clusters:
\begin{enumerate}
  \item \label{step:gabriel-cluster}
    \textbf{Cluster:}
    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
    assignment rule $\hGY : \mathbb{R}^q \to \{ 1, \dotsc, k \}$ and the
    clusters means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
    be the assigned cluster for row $i$.

  \item \label{step:gabriel-classify}
    \textbf{Classify:}
    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
    $\hGX : \mathbb{R}^p \to \{ 1, \dotsc, k \}$.

@@

Because K-means algorithm is a key part in the proof below, it's convenient to have a review of K-means here:\\
Given a set of observations $\mathbf{X} = (x_1,...,x_n)$, and the number of clusters $K$, K-means' goal is to minimize the within cluster dispersion. That is, find the partition $\mathbb{P}_K$ of $\mathbf{X}$, where $\mathbb{P}_K(\mathbf{X})= (p_1,p_2,...,p_K)$, such that
\[	\underset{\mathbb{P}_K}{\operatorname{argmin}} \sum^K_{i=1}\sum_{x \in p_i } ||x-\hat{\mu}_i||^2\]
where $\hat{\mu}_i$ is the sample mean of $p_i$.
\subsection{Self-Consistency}
Same as in section $2$, let the data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been partitioned as 
 \[ 
 \begin{pmatrix}
\mathbf{X}^L_{train} & \mathbf{X}^R_{train}\\
  \mathbf{X}^L_{test}  & \mathbf{X}^R_{test}\\
  \end{pmatrix}
 \]
\subsection*{Assumptions}
\begin{enumerate}[label={Assumption},leftmargin=*]
\item $1.$ There are $K$ true centers $(\mu_1,...,\mu_K)$ where $\mathbf{X}$ generated from
\item $2.$ The $K$ partitioned centers $(\mu^L_1,...,\mu^L_K)$ are all distinct in $L$ dimensions: $||\mu^L_i-\mu^L_j|| \neq 0$ for any $i \neq j$; $(\mu^R_1,...,\mu^R_K)$ are all distinct in $R$ dimensions
\item $3.$ The training set contains at least one member of each cluster
\item $4.$ The test set contains at least one member of each cluster
\item $5.$ There is no noise, so that $$x_i = \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\} $$ and $$\mu \left(c_i\right) = \mu_j \hspace{0.2in} \text{iff} \hspace{0.2in} c_i = j$$ where $c_i$ is the true cluster for observation $i$, which implies $$x^L_i = \mu^L\left(c_i\right) \in \{\mu^L_1,...,\mu^L_K\},\hspace{0.2in} x^R_i = \mu^R\left(c_i\right) \in \{\mu^R_1,...,\mu^R_K\}$$
\end{enumerate}

@@

\subsection{Single Cluster With Noise}

Our proof relies on results from \citet{pollard1981strong}, who analyzed
$k$-means and showed that the within-cluster sum of squares converges to a
population quantity.  Specifically, for any finite set $A \subset
\mathbb{R}^d$ and probability measure $P$ supported on $\mathbb{R}^d$, define
the population within-cluster sum of squares as
\[
  W(A, P) = \int \min_{a \in A} \|x - a\|^2 P(dx).
\]
Define
\(
  \sA_k
    =
    \{ A \subset \mathbb{R}^d : \text{$A$ contains $k$ or fewer points} \}
\)
and set
\(
  m_k(P) = \inf_{A \in \sA_k}  W(A, P). % : \text{$A$ contains $k$ or fewer points} \}.
\)
Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and if $P_n$
is the empirical measure formed by placing mass $1/n$ on each $X_i$.  Let
$A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for $A_n$ to
converge to a minimizer of $W(\cdot, P)$.  His main theorem assumes that the
minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
(this fails, for example, with a bivariate normal distribution with identity
covariance whenever $k \geq 2$).  It is straightforward to generalize his
result to cases with multiple maximizers.

To state the consistency theorem define
\[
  \sbA_k = \{ \bar A \subseteq \mathbb{R}^d
            : W(\bar A, P) = m_k(P)
              \text{ and $\bar A$ contains $k$ or fewer points} \}.
\]
We will later show that $\sbA_k$ is nonempty and compact whenever
$\int \|x\|^2 P(dx) < \infty$.
Let $H(\cdot, \cdot)$ denote the Hausdorff metric, defined for compact subsets
$A, B$ of $\mathbb{R}^d$ by $H(A,B) < \delta$ if and only if every point of
$A$ is within Euclidean distance $\delta$ of at least one point of $B$, and
vice versa.  We can now state the generalized version of Pollard's $k$-means
consistency theorem.

@@

We first show that if $M$ is large enough, then there exists a set $B \in \sE_k$
and a positive $\varepsilon$ with $W(B,P) < m_{k-1}(P) - \varepsilon$.  This
will establish that $m_{k}(P) < m_{k-1}(P)$, and the set $B$ will have further
utility later in the proof.
Take $\bar A_{k-1} \in \sE_{k-1}$ to be a set with at most $k-1$ points satisfying
$m_{k-1}(P) = W(\bar A_{k-1},P)$.
Since the support of $P$ contains least $k$ points, there exists a point~$b_1$ not
in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is large
enough so that $b_1 \in B(5M)$.  Take $B \in \sE_k$ to be the set of $k$
points gotten by adjoining $b_1$ to $\bar A_{k-1}$.  Let
\(
  \delta = \min_{a \in \bar A_{k-1}} \| b_1 - a \|
\)
and set $R = \{ x \in \mathbb{R}^d : \|x - b_1\| < \delta/3 \}$; since $b_1$
is in the support of $P$, the neighborhood $R$ has positive $P$~measure.
Further,
\[
  \int_{R} \min_{a \in \bar A_{k-1}} \|x - a\|^2 P(dx)
  \geq
  \int_{R} \min_{a \in \bar A_{k-1}} (\|b_1 - a\| - \delta/3)^2 P(dx)
  = (4/9) \delta^2 P(R),
\]
and
\[
  \int_{R} \min_{b \in B} \|x - b\|^2 P(dx)
  =
  \int_{R} \|x - b_1\|^2 P(dx)
  \leq
  (1/9) \delta^2 P(R).
\]
Thus, if
\(
  \varepsilon < (3/9) \delta^2 P(R)
\)
then
\(
  W(B,P) \leq W(\bar A_{k-1},P) - (3/9) \delta^2 P(R)
         < m_{k-1}(P) - \varepsilon.
\)

@@

\begin{lemma}
\label{lem:pollard-compact}
Suppose that $\int \|x\|^2 P(dx) < \infty$, that the support of $P$ has at
least $k$ points, and that $k > 1$.  Let $A_n$ minimize $W(\cdot, P_n)$ over
$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If
$W(B_n, P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then
there exists a finite $M$ such that as if $n$ is large enough,
then, almost surely, $A_n$ is in the class of sets
\(
  \sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer points} \},
\)
where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
origin.
\end{lemma}

@@

\begin{proof}[Proof of Theorem~\ref{lem:kmeans-consistency}]

The proof proceeds inductively over $k$.  For $k=1$, the result follows from
the strong law of large numbers.  For $k > 1$, assume that the result holds
for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
then the result is trivially true.  Otherwise, the induction hypothesis
implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains
$\sbA_k$.

The set $\sE_k$ is compact and the map $f: \sE_k \to \mathbb{R}$ with $f(A) =
W(A, P)$ is continuous.  Thus, for every $\varepsilon > 0$ there exists an
$\eta > 0$ depending on $\varepsilon$ such that if $A \in \sE_k$ and $H(A,
\bar A) \geq \varepsilon$ for every $\bar A \in \sbA_k$, then $W(A,P) \geq
m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
$W(A_n, P_n) \leq W(\bar A, P_n)$.  Lemma~\ref{lem:kmeans-uniform-slln} ensures
that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large enough,
$W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an $\bar A_n
\in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.
\end{proof}

Now, define $\mathcal{\bar A} = \mathcal{\bar A}(k)$ to be the class of sets
$\bar A$ with $k$ or fewer points such that $W(\bar A,P) = m_k(P)$.
Lemmas~\ref{lem:pollard-compact} and~\ref{lem:kmeans-continuous} together
with the extreme value theorem imply that this class is nonempty.

To prove
this, it suffices to analyze a single fold of the cross-validation procedure.
We will assume the following setup:

@@

@@

Same with previous, the data set $\mathbf{X} \in \mathbb{R}^{n \times p}$ been
partitioned as $\mathbf{X}^L \in \mathbb{R}^{n \times L}$ and $\mathbf{X}^R
\in \mathbb{R}^{n \times R}$, with following set of assumptions:

>>>>

\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\dataX}{\mathfrak{X}}
\newcommand{\Xtrain}{X_{\text{train}}}
\newcommand{\Ytrain}{Y_{\text{train}}}
\newcommand{\Xtest}{X_{\text{test}}}
\newcommand{\Ytest}{Y_{\text{test}}}
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}
\newcommand{\bpi}{\bar \pi}
\newcommand{\bmuX}{\bar \mu^{X}}
\newcommand{\bmuY}{\bar \mu^{Y}}
\newcommand{\sA}{\mathcal{A}}
\newcommand{\sbA}{\mathcal{\bar A}}
\newcommand{\sC}{\mathcal{C}}
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sN}{\mathcal{N}}
\newcommand{\sX}{\mathcal{X}}

@@

Cross-validation is commonly used for model selection in supervised learning
problems.  In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
cross-validation, we permute the rows and columns of the data matrix and then
partition the rows and columns as $N = n + m$ and $P = p + q$ for 
non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
columns as ``predictors'' and the last $q$ columns as ``responses'';
similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
as ``test''.  In block form, the permuted data matrix is
\[
  \dataX
  =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix},
\]
where
$\Xtrain \in \R^{n \times p}$,
$\Ytrain \in \R^{n \times q}$,
$\Xtest \in  \R^{m \times p}$,
and
$\Ytest \in  \R^{m \times q}$.

Given such a partition of $\dataX$, we perform four steps for each value of
$k$, the number of clusters:
\begin{enumerate}
  \item \label{step:gabriel-cluster}
    \textbf{Cluster:}
    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
    clusters means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
    be the assigned cluster for row $i$.

  \item \label{step:gabriel-classify}
    \textbf{Classify:}
    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.

@@

Because K-means algorithm is a key part in the proof below, it's convenient to have a review of K-means here:\\
Given a set of observations $\mathbf{X} = (x_1,...,x_n)$, and the number of clusters $K$, K-means' goal is to minimize the within cluster dispersion. That is, find the partition $\mathbb{P}_K$ of $\mathbf{X}$, where $\mathbb{P}_K(\mathbf{X})= (p_1,p_2,...,p_K)$, such that
\[	\underset{\mathbb{P}_K}{\operatorname{argmin}} \sum^K_{i=1}\sum_{x \in p_i } ||x-\hat{\mu}_i||^2\]
where $\hat{\mu}_i$ is the sample mean of $p_i$.
\subsection{Self-Consistency}
Same as in section $2$, let the data set $\mathbf{X} \in \R^{n \times p}$ been partitioned as
 \[ 
 \begin{pmatrix}
\mathbf{X}^L_{train} & \mathbf{X}^R_{train}\\
  \mathbf{X}^L_{test}  & \mathbf{X}^R_{test}\\
  \end{pmatrix}
 \]
\subsection*{Assumptions}
\begin{enumerate}[label={Assumption},leftmargin=*]
\item $1.$ There are $K$ true centers $(\mu_1,...,\mu_K)$ where $\mathbf{X}$ generated from
\item $2.$ The $K$ partitioned centers $(\mu^L_1,...,\mu^L_K)$ are all distinct in $L$ dimensions: $||\mu^L_i-\mu^L_j|| \neq 0$ for any $i \neq j$; $(\mu^R_1,...,\mu^R_K)$ are all distinct in $R$ dimensions
\item $3.$ The training set contains at least one member of each cluster
\item $4.$ The test set contains at least one member of each cluster
\item $5.$ There is no noise, so that $$x_i = \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\} $$ and $$\mu \left(c_i\right) = \mu_j \hspace{0.2in} \text{iff} \hspace{0.2in} c_i = j$$ where $c_i$ is the true cluster for observation $i$, which implies $$x^L_i = \mu^L\left(c_i\right) \in \{\mu^L_1,...,\mu^L_K\},\hspace{0.2in} x^R_i = \mu^R\left(c_i\right) \in \{\mu^R_1,...,\mu^R_K\}$$
\end{enumerate}

@@

\clearpage

\section{Asymptotic Analysis of $k$-Means}

To analyze the asymptotic properties of of Gabriel cross validation, we first
need to understand the properties of the $k$-means procedure.  We will use
results from \citet{pollard1981strong}, who analyzed $k$-means and showed that
the within-cluster sum of squares converges to a population quantity.  To
introduce Pollard's result, for any finite set $A \subseteq \R^d$ and
probability measure $P$ supported on $\R^d$, define the population
within-cluster sum of squares as
\[
  W(A, P) = \int \min_{a \in A} \|x - a\|^2 P(dx).
\]
Define
\(
  \sA_k
    =
    \{ A \subseteq \R^d : \text{$A$ contains $k$ or fewer points} \}
\)
and set
\(
  m_k(P) = \inf_{A \in \sA_k}  W(A, P).
\)
Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and that $P_n$
is the empirical measure formed by placing mass $1/n$ on each $X_i$.  Let
$A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for $A_n$ to
converge to a minimizer of $W(\cdot, P)$.  His main theorem assumes that the
minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
(this fails, for example, with a bivariate normal distribution with identity
covariance whenever $k \geq 2$).  It is straightforward to generalize his
result to cases with multiple maximizers.

To state the generalized version of Pollard's consistency theorem, define
\[
  \sbA_k = \{ \bar A \subseteq \R^d
            : W(\bar A, P) = m_k(P)
              \text{ and $\bar A$ contains $k$ or fewer points} \}.
\]
We will later show that $\sbA_k$ is nonempty and compact whenever $\int
\|x\|^2 P(dx) < \infty$.  Let $H(\cdot, \cdot)$ denote the Hausdorff metric,
defined for compact subsets $A, B$ of $\R^d$ by $H(A,B) < \delta$ if
and only if every point of $A$ is within Euclidean distance $\delta$ of at
least one point of $B$, and vice versa.  We can now state the generalized
$k$-means consistency theorem.

@@

We first show that if $M$ is large enough, then there exists a set $B \in
\sE_k$ and a positive $\varepsilon$ with $W(B,P) < m_{k-1}(P) - \varepsilon$.
This will establish that $m_{k}(P) < m_{k-1}(P)$, and the set $B$ will have
further utility later in the proof.  Take $\bar A_{k-1} \in \sE_{k-1}$ to be a
set with at most $k-1$ points satisfying $m_{k-1}(P) = W(\bar A_{k-1},P)$.
Since the support of $P$ contains least $k$ points, there exists a point~$b_1$
not in $\bar A_{k-1}$ that lies in the support of $P$.  Ensure that $M$ is
large enough so that $b_1 \in B(5M)$.  Take $B \in \sE_k$ to be the set of $k$
points gotten by adjoining $b_1$ to $\bar A_{k-1}$.  Let
\(
  \delta = \min_{a \in \bar A_{k-1}} \| b_1 - a \|
\)
and set $R = \{ x \in \R^d : \|x - b_1\| < \delta/3 \}$; since $b_1$
is in the support of $P$, the neighborhood $R$ has positive $P$~measure.
Further,
\[
  \int_{R} \min_{a \in \bar A_{k-1}} \|x - a\|^2 P(dx)
    \geq
      \int_{R} \min_{a \in \bar A_{k-1}} (\|b_1 - a\| - \delta/3)^2 P(dx)
    =
      (4/9) \delta^2 P(R),
\]
and
\[
  \int_{R} \min_{b \in B} \|x - b\|^2 P(dx)
    =
      \int_{R} \|x - b_1\|^2 P(dx)
    \leq
      (1/9) \delta^2 P(R).
\]
Thus, if
\(
  \varepsilon < (3/9) \delta^2 P(R)
\)
then
\(
  W(B,P)
    \leq
      W(\bar A_{k-1},P) - (3/9) \delta^2 P(R)
    <
      m_{k-1}(P) - \varepsilon.
\)

@@

\begin{lemma}
\label{lem:pollard-compact}
Suppose that $\int \|x\|^2 P(dx) < \infty$, that the support of $P$ has at
least $k$ points, and that $k > 1$.  Let $A_n$ minimize $W(\cdot, P_n)$ over
$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If $W(B_n,
P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then there exists a finite
$M$ such that as if $n$ is large enough, then, almost surely, $A_n$ is in the
class of sets
\(
  \sE_k = \{ A \subseteq B(5M) : \text{$A$ contains $k$ or fewer points} \},
\)
where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
origin.
\end{lemma}

@@

\begin{proof}[Proof of Theorem~\ref{lem:kmeans-consistency}]
The proof proceeds inductively over $k$.  For $k=1$, the result follows from
the strong law of large numbers.  For $k > 1$, assume that the result holds
for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
then the result is trivially true.  Otherwise, the induction hypothesis
implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains
$\sbA_k$.

The set $\sE_k$ is compact and the map $f: \sE_k \to \R$ with $f(A) =
W(A, P)$ is continuous.  Thus, for every $\varepsilon > 0$ there exists an
$\eta > 0$ depending on $\varepsilon$ such that if $A \in \sE_k$ and $H(A,
\bar A) \geq \varepsilon$ for every $\bar A \in \sbA_k$, then $W(A,P) \geq
m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
$W(A_n, P_n) \leq W(\bar A_0, P_n)$.  Lemma~\ref{lem:kmeans-uniform-slln}
ensures that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large
enough, $W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an
$\bar A_n \in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.
\end{proof}

The following lemma is not required for the $k$-means consistency theorem, but
it will prove necessary for analyzing the asymptotic behavior of Gabriel
cross-validation.

@@

\section{Asymptotic Analysis of Gabriel Cross-Validation}

To prove consistency of Gabriel cross-validation, it suffices to analyze a
single fold of the cross-validation procedure.  We will assume the following
setup:

\clearpage

@@

Same with previous, the data set $\mathbf{X} \in \R^{n \times p}$ been
partitioned as $\mathbf{X}^L \in \R^{n \times L}$ and $\mathbf{X}^R
\in \R^{n \times R}$, with following set of assumptions:
